---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
Unzip the data file.
```{r}
setwd("/Users/enogrob/TODAY/Reproducible_Research/Week2/RepData_PeerAssessment1")
unzip("activity.zip")
```

Read the CSV file. Convert the data frame to a data table using the data.table package.
```{r eval=FALSE}
install.packages("data.table")
```
```{r}
library(data.table)
dt <- read.csv(file.path(getwd(), "activity.csv"))
dt <- data.table(dt)
```

Verify that the number of rows in the dataset is the expected value of 17,568.
```{r}
check <- nrow(dt) == 17568
if (check == FALSE) stop("The number of rows in the dataset is not 17,568.")
```

Convert the date variable to a date class. And look at the structure of the dataset.
```{r}
dt <- dt[, date := as.Date(date)]
setkey(dt, date, interval)
str(dt)
dt
```

## What is mean total number of steps taken per day?
Aggregate the number of steps taken each day. Days with missing values (NA) will have NA when aggregated.
```{r}
dtDaily <- dt[, list(sumSteps = sum(steps)), date]
head(dtDaily)
```

Plot a histogram of the total number of steps taken each day.
```{r eval=FALSE}
install.packages("ggplot2")
```
```{r}
library(ggplot2)
ggplot(dtDaily, aes(x=sumSteps)) +
    geom_histogram(alpha=1/2, binwidth=1000)
```

Calculate the mean and median total number of steps taken per day before imputing.
```{r eval=FALSE}
install.packages("xtable")
```
```{r}
library(xtable)
tab <- dtDaily[, list(n = .N, nValid = sum(!is.na(sumSteps)), mean = mean(sumSteps, na.rm=TRUE), median = median(sumSteps, na.rm=TRUE))]
print(xtable(tab), type="html", include.rownames=FALSE)
```

Copy the data table dtDaily before imputation to be used later.
```{r}
dtDaily <- dtDaily[, status := "Before imputation"]
dtDailyBeforeImputation <- dtDaily
```

## What is the average daily activity pattern?
Aggregate the average number of steps taken by 5-minute interval.
```{r}
dtIntervals <- dt[, list(meanSteps = mean(steps, na.rm=TRUE)), interval]
```

Plot a time series of the 5-minute interval and the average number of steps taken across all days.
```{r}
ggplot(dtIntervals, aes(x=interval, y=meanSteps)) +
    geom_line()
```

## Imputing missing values
Calculate the total number of missing values.
```{r}
dt <- dt[, isStepsMissing := is.na(steps)]
tab <- dt[, .N, isStepsMissing]
print(xtable(tab), type="html", include.rownames=FALSE)
```

Use the VIM package to impute missing values of the steps variable. 
```{r eval=FALSE}
install.packages("VIM")
```
```{r}
library(VIM)
dt <- kNN(dt)
```

The kNN function returns a dataset with all NAs replaced. So the steps variable now contains imputed values replacing the NAs.
Verify that there are no missing values for steps after imputation.
```{r}
tab <- dt[, .N, list(isMissing = is.na(steps))]
print(xtable(tab), type="html", include.rownames=FALSE)
```

Verify that missingness is complete for an entire day. Show all days with at least 1 missing value for the steps variable. Calculate the proportion of records with missing values for each such day. All proportions are 100%.
```{r}
dtMissingness <- dt[, list(countMissing = sum(isStepsMissing), countRecords = .N, propMissing = sum(isStepsMissing / .N)), date]
dtMissingness[countMissing > 0]
```

### After imputation of missing values

Aggregate the number of steps taken each day.
```{r}
dtDaily <- dt[, list(sumSteps = sum(steps), isImputed = sum(steps_imp) > 0), date]
head(dtDaily)
```

Plot a histogram of the total number of steps taken each day after imputing and compare with the histogram before imputing. Need to add an isImputed column to dtDailyBeforeImputation to make rbind work.
```{r}
dtDaily <- dtDaily[, status := "After imputation"]
dtDailyBeforeImputation <- dtDailyBeforeImputation[, isImputed := FALSE]
dtDaily <- rbind(dtDaily, dtDailyBeforeImputation, use.names=TRUE)
ggplot(dtDaily, aes(x=sumSteps, fill=isImputed)) +
    geom_histogram(alpha=1/2, binwidth=1000) +
  facet_wrap(~ status, nrow=2) +
  theme(legend.position="bottom")
```

Calculate the mean and median total number of steps taken per day after imputing.
```{r}
tab <- dtDaily[, list(n = .N, nValid = sum(!is.na(sumSteps)), mean = mean(sumSteps, na.rm=TRUE), median = median(sumSteps, na.rm=TRUE)), status]
print(xtable(tab), type="html", include.rownames=FALSE)
```

The median of the imputed values is the same as the original values where missing values were not imputed. However, the mean of the imputed values is less than the original values. The overall impact of the imputed values is to lower the estimates of the number of steps taken each day.

## Are there differences in activity patterns between weekdays and weekends?
Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day. Use this solution to collapse the factor values for day of week. Verify that dayOfWeek and dayType are factor class variables.
```{r}
levels <- c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")
newLevels <- c("Weekend", rep("Weekday", 5), "Weekend")
dt <- dt[, dayOfWeek := factor(weekdays(date), levels=levels)]
dt <- dt[, dayType := factor(newLevels[dayOfWeek])]
dt[, .N, list(dayType, dayOfWeek)]
message(sprintf("Is dayOfWeek a factor? %s. Is dayType a factor? %s", is.factor(dt$dayOfWeek), is.factor(dt$dayType)))
```

Aggregate the average number of steps taken by 5-minute interval. Use the imputed values in the steps variable.
```{r}
dtIntervals <- dt[, list(meanSteps = mean(steps, na.rm=TRUE)), list(dayType, interval)]
Plot two time series (one for weekdays and the other for weekends) of the 5-minute intervals and average number of steps taken (imputed values).

ggplot(dtIntervals, aes(x=interval, y=meanSteps, color=dayType)) +
    geom_line() +
    facet_wrap(~ dayType, nrow=2) +
    theme(legend.position="none")
```

It's a bit hard to discern any differences. So overlay the time series on a single plot instead of using a panel plot.
```{r}
ggplot(dtIntervals, aes(x=interval, y=meanSteps, color=dayType)) +
    geom_line() +
    theme(legend.position="bottom")
```